---
title: "Creating a map of the US and its territories"
format: html
---

This scrips walks through how to plot the a map of the continental US with additional insets of US territories. For this example we use the `shift_geometry()` function from the `tigris` package to plot the continental US, Alaska, Hawaii and Puerto Rico in one map.

The idea of this script is to present an example of how to work with spatial data in R using the `sf` package and introduce tools and techniques that will help you create the maps you like for your project.

## Set up
We will start by loading the packages we will be using to create our map.

```{r}
library(tigris)
library(sf)
library(dplyr)
library(ggplot2)
library(janitor)
library(readr)
library(stringr)
library(ggrepel)
```


## Read data
Next step is to read the data we are going to be using today. In this case we will use 4 data sources.

- The US state boundaries that we can load directly from the `tigris` package.
- The continental US EPA Ecoregions used in the National Aquatic Resource Surveys.
- The Level 3 EPA Ecoregions by state for the state od Alaska
- List of states and their corresponding CASC region (this data set was manually created, see us-casc-regions.R script)



### 1. Load US map from `tigris`

```{r}
## read data
# cb = T + 20m resol, removes American Samoa, Maraina Islands, Guam and Virgin Islands
us_states <- states(cb = TRUE, 
                    resolution = "20m")

## shift geometries for easy plotting
us_states_shift <- us_states %>% 
  shift_geometry()

```


### 2. Read in Ecoregions

**a) Continental US**

For the continental US we are using the aggregated ecoregions used in the National Aquatic Resource Surveys. You can download this data at the [EPA website](https://www.epa.gov/national-aquatic-resource-surveys/ecoregions-used-national-aquatic-resource-surveys).

These data divided the continental US in 9 ecoregions. 


```{r}
## read data
ecoregions_cont_us <- read_sf("data/Aggr_Ecoregions_2015.shp")

unique(ecoregions_cont_us$WSA9_NAME)
st_crs(ecoregions_cont_us) #Albers, NAD83

```


**b) Alaska L3 Ecoregions**

To add additional territories, we will have to download individual files for each area. Here an example for Alaska. In this case we are downloading the shapefiles for [Level III EPA ecoregions for Alaska](https://www.epa.gov/eco-research/ecoregion-download-files-state-region-10#pane-01).

```{r}
## Read data
ecoregion_ak_l3 <- read_sf("data/ak_eco_l3.shp")
st_crs(ecoregion_ak_l3) # EPSG",3338, Alaska Albers
plot(ecoregion_ak_l3$geometry)

```


### 3. US major cities

Shapefile indicating the locatio of the major cities across the US. Data downloaded form [here](https://hub.arcgis.com/datasets/esri::usa-major-cities/explore?location=11.509491%2C76.498957%2C3.25)


```{r}
## read
us_cities <- read_sf("data/USA_Major_Cities.shp")
st_crs(us_cities) #WGS 84 / Pseudo-Mercator
plot(us_cities$geometry)

## transform to desired crs
us_cities_nad83 <- st_transform(us_cities,
                                crs = st_crs(us_states))
st_crs(us_cities_nad83)
plot(us_cities_nad83$geometry)

## shift geometries for easy plotting
us_cities_shift <- us_cities_nad83 %>% 
  shift_geometry()

plot(us_cities_shift$geometry)

```


#### 4. States by CASC region
```{r}
## read data
us_state_casc <- read_csv("data/state_by_casc_region.csv")

## merge polygons to get CASC region areas
casc_shp <- us_states %>% 
  left_join(us_state_casc, by= c("GEOID", "STUSPS", "NAME")) %>% 
  group_by(casc_region) %>% 
  summarise(geometry = st_union(geometry))
plot(casc_shp$geometry)

## shift geometries for easy plotting
casc_shift <- casc_shp %>% 
  shift_geometry()

plot(casc_shift$geometry)

```



## Cleaning Ecoregion data
To be able to plot all the regions in one map, we need to get both ecoregions files in the same format. This means column names should ,atch so we can bind both files.

```{r}
## Rename columns to a generic name
ecoregions_cont_us_clean <- ecoregions_cont_us %>% 
  rename(code = WSA9,
         name = WSA9_NAME)
  
## Select two main colums from Alaska ecoregion file and rename to match continental file + transforming to the smae CRS than continental file.
## note: choosing L1 because it is the mos broad ecoregion (4 ecoregions for alaska)
ecoregion_ak_clean <- ecoregion_ak_l3 %>% 
  select(code = NA_L1CODE,
         name = NA_L1NAME) %>% 
  mutate(name = str_to_sentence(name)) %>% 
  st_transform(st_crs(ecoregions_cont_us_clean))

##checking outcomes
unique(ecoregion_ak_clean$name)
st_crs(ecoregion_ak_clean)
plot(ecoregion_ak_clean$geometry)

```

## Bind ecoregions data into one file

```{r}
ecoregions_all <- bind_rows(ecoregions_cont_us_clean, 
                            ecoregion_ak_clean)


# shift geom
ecoregions_shift <- ecoregions_all %>% 
  shift_geometry()

plot(ecoregions_shift$geometry)
```


## Plot ecoregions + cities + casc

```{r}

us_plot <-  ggplot()+
  geom_sf(data = ecoregions_shift,
          aes(fill = name))+
  geom_sf(data = us_states_shift,
          fill = NA,
          color = "darkgray")+
  geom_sf(data = us_cities_shift)+
  geom_sf(data = casc_shift,
          color = "red",
          fill = NA)+
  theme_void()+
  scale_fill_viridis_d()

us_plot
```


## Customize map
```{r}

finalized_plot <-  ggplot()+
  geom_sf(data = ecoregions_shift,
          aes(fill = name),
          alpha = 0.5)+
  geom_sf(data = us_states_shift,
          fill = NA,
          color = "darkgray")+
  geom_sf(data = us_cities_shift,
          alpha = 0.3)+
  geom_sf(data = casc_shift,
          color = "black",
          size = 6,
          fill = NA)+
  #an alternative to label your can use geom_sf_text, plots just the text not the rectagle aroud it.
  geom_sf_label(data = casc_shift[c(1:7), ],
               aes(label = casc_region),
               size = 3)+
  geom_label_repel(
    data = casc_shift[c(8:10), ],
    aes(label = casc_region, 
        geometry = geometry),
    stat = "sf_coordinates",
    size = 3,
    min.segment.length = 2)+
    # colour = "magenta",
    # segment.colour = "magenta")+
  theme_void()+
  scale_fill_viridis_d(name = "Ecoregion")+
  labs(title = "Cities across Ecoregions in the US",
       subtitle = "Map is devided into CASC regions")+
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))


finalized_plot

```


